---
title: "Getting started with rcol"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rcol}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`rcol` wraps the [ChecklistBank](https://www.checklistbank.org) API to give R
users idiomatic access to the [Catalogue of
Life](https://www.catalogueoflife.org) (COL) and every other public dataset in
ChecklistBank.

```{r setup}
library(rcol)
```

## Choosing a dataset

Almost every function takes a `dataset =` argument identifying which checklist
to work against. You can pass:

* a numeric dataset key, e.g. `3` (the COL project),
* a dataset alias such as `"COL25"` (the 2025 annual release), or
* one of the *magic* COL release aliases that always resolve to the most recent
  release.

The COL is published in two cadences (monthly and annual) and two flavours: a
**base** release and an **extended** release (`XR`) that merges in additional
data. `clb_col_release()` returns the right identifier:

```{r}
clb_col_release("monthly")                  # "3LR"   latest monthly base release
clb_col_release("monthly", extended = TRUE) # "3LXR"  latest monthly extended release
clb_col_release("annual")                   # e.g. "COL25"

# Every release, e.g. to find historical annual editions
clb_col_releases()
```

The package default is `"3LXR"`, the latest monthly extended release.

## Matching names

`clb_match()` returns the single best match — the COL analogue of GBIF's name
backbone matching:

```{r}
clb_match("Panthera leo")
clb_match("Abies alba", rank = "species", code = "botanical")
```

Use `clb_match_verbose()` to see all candidate usages, and
`clb_match_checklist()` to match a whole vector or data frame in parallel:

```{r}
clb_match_verbose("Oenanthe")

clb_match_checklist(c("Panthera leo", "Bufo bufo", "Abies alba"))
```

## Parsing

The ChecklistBank parsers are exposed too — the scientific name parser and all
the controlled-value parsers:

```{r}
clb_parse_name("Abies alba Mill.")

clb_parsers()
clb_parse("rank", c("spec", "fam.", "ssp"))
```

## Datasets

```{r}
clb_dataset_search("mammal")
clb_dataset("3LXR")
clb_dataset_metrics("COL25")
```

## Taxa and the tree

```{r}
# Look up a usage and its context
clb_usage("4CGXP", dataset = "3LR")
clb_classification("4CGXP", dataset = "3LR")
clb_synonyms("6DBT", dataset = "3LR")
clb_vernacular(id = "4CGXP", dataset = "3LR")

# Navigate the tree
roots <- clb_tree(dataset = "3LR")
clb_children(roots$id[1], dataset = "3LR")

# Full-text search
clb_usage_search("Felidae", dataset = "3LR")
clb_suggest("Panth", dataset = "3LR")
```

## Pointing at another server

Set `CLB_BASE_URL` to target the development API, or pass `server =` to
`clb_match()` to match against a locally running
[dockerized matching container](https://www.catalogueoflife.org/tools/matching):

```{r}
Sys.setenv(CLB_BASE_URL = "https://api.dev.checklistbank.org")

# docker run -p 8080:8080 docker.gbif.org/matching-ws:xcol-amd64-latest
clb_match("Panthera leo", server = "http://localhost:8080")
```
